---
title: "Class 9: Candy Mini Project"
author: Jolie Adair (PID A17337497)
format: pdf
---

## Background

In this mini-project, you will explore FiveThirtyEight's Halloween Candy Dataset.

We will use lots of **ggplot** some basic stats, correalation analysis and PCA to make sense of the landscape of US candy - something hopefully more relocatable than the prote 
transcriptomics work that we will use these methods throughout the rest of the course.

## Data Import 

peak with `head()`

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```


> Q1. How many different candy types are in this dataset?

There are 85 different types of candy in this dataset.

```{r}
nrow(candy)
```


> Q2. How many fruity candy types are in the dataset?

```{r}
candy$fruity
```

```{r}
sum(candy$fruity)

```


> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Hershey's Milk Chocolate") |> 
  select(winpercent)
```



> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", "winpercent"]
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, if we did PCA on the data, winpercent would dominate the scale.

```{r}
library("skimr")
skim(candy)
```


> Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
library("skimr")
skim(candy$chocolate)
```


## Exploratory Analysis

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```
```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram()
```


> Q9. Is the distribution of winpercent values symmetrical?

No the distribution is not symmertical it is slightly skewed to left.

> Q10. Is the center of the distribution above or below 50%?



```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate candy is higher ranked than fruit candy.

```{r}
choc.candy <-candy[candy$chocolate ==1, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```


```{r}
fruity.win <- candy[candy$fruity == 1,]$winpercent
mean(fruity.win)
```


> Q12. Is this difference statistically significant?

Chocolate has a smaller p-value than Fruity candy

]

```{r}
chocolate_rank <- candy$winpercent[as.logical(candy$chocolate)]
candy_rank <- candy$winpercent[as.logical(candy$fruity)]

t.test(x = chocolate_rank, y = candy_rank)
```

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
y = c("y", "a", "z")
sort(y)
```

```{r}
y
order(y)
```
```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind, ])
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[ord.ind, ], 5)
```


> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}

ggplot(candy) + 
  aes(winpercent,
      reorder(row.names(candy),winpercent)) +
  geom_col() +
  ylab("")
```

```{r}
ggplot(candy) + 
  aes(winpercent,
      reorder(row.names(candy),winpercent), 
      fill=chocolate) +
  geom_col() +
  ylab("")
```
We need a custom color vector

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
my_cols <- rep("black", nrow(candy))
#my_cols[1] <- "red"  first element is red
my_cols[candy$chocolate==1] <- "chocolate"
my_cols[candy$bar==1] <- "darkgreen"
my_cols[candy$fruity==1] <- "hotpink"
my_cols
```
```{r}
ggplot(candy) + 
  aes(winpercent,
      reorder(row.names(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

```


> Q19. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

Nik L Nip

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Twix
Kit Kat
Reese's Peanut Butter Cup
starburst
Reese's Minaiatures


## Exploring the correlation structure 

```{r}
cij <- cor(candy)
```

```{r}
library(corrplot)
corrplot(cij)
```


> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruit are anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar are the most positively correlated.

##  Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Score Plot...
```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=my_cols)+
  geom_text_repel(max.overlaps = 5, col=my_cols)

```

```{r}
ggplot(pca$rotation) + 
  aes(PC1, 
      reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```

```{r}
#library(plotly)
#ggplotly(p)
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

PC1 is driven positively due to fruity and hard candies while chocolate and bar candies make it negative.


> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

The combined characteristics that make a winning candy is having the characteristics of chocolate and a bar, they have higher winpercent values and positive correlation as seen on the plots.