---
title: "Class 8: Cancer Mini Project"
author: "Jolie (PID A17337497)"
format: pdf
---

## Background

In today;s class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer FNA biopsy data set.

## Data Import

We start by importing our data. It is a CSV file so we will use the `read.csv()` function.

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names =1)
```

have a wee peak at the first few rows"

```{r}
head(wisc.df)
```

Make sure to remove the first `diagonsis` column - I don't want to use this for my machine learning models. We will use it later to compare our results to the expert diagnosis.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Exploratory Data Analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```


> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```

```{r}
table(wisc.df$diagnosis)
```


> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)
length(grep("_mean$", colnames(wisc.data)))
```



## Principal Component Analysis

The main functions here is `prcomp()` and we want to make sure we set the optional argument `scale=TRUE` :


```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
summary(wisc.pr)
```

 Pc2 = 0.1897, there are getting smaller and smaller.

If you add up all the PCs..., the cumulative of 2 PCs is 0.6324 (63%)

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

PC1 captures 0.4427, 44%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs is 72% 

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs is 91% 

Our main PCA "score plot" or "PC plot" of results:
```{r}
library(ggplot2)
```
```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```


> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

Each point is a person, M = malignant    B = Benign     
- similar characteristics will cluster together
- Cell PC1 score (cell1 # influence on PC1 #)+()
- low weight PCs don't contribute/influence the cells much compared to the higher weighted PCs

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots? - I notice that they are more towards the bottom of the graph. 

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```


Scree-plot: 
 
```{r}
library(factoextra)
```


```{r}
fviz_eig(wisc.pr, addlabels= TRUE)
```


```{r}
ggplot(wisc.pr$rotation)+
  aes(PC1, 
      order( rownames(wisc.pr$rotation), PC1)) +
  geom_col()
```


Collectively these two plot ("score plot" and "loading plot") tells us that 
if cells nucleus are deeply indented ("conclave"), irregular circular ("compactness"), and have large "perimeter" values they tend to be malignant...

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one? 

```{r}
wisc.pr$rotation["concave.points_mean",1]
```


## Hierarchical Clustering

First scale the data (with the `scale()` function), then calculate a distance matrix (with the `dist()` function).  Then cluster with the `hclust()` function and plot"

```{r}
wisc.hclust <- hclust(dist(scale(wisc.data)))
```


> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=20, col="red")
```

You can also use the `cutree()` function with a arugment `k=4` rather than `h=height`

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4) 
table(wisc.hclust.clusters)
```


## Combining methods

Here we will take our PCA results and use those input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principal component axis/variables)  and use a subset of the PCs that capture the most variance as input for `hclust()`

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method="ward.D2")
plot(wisc.pr.hclust)
```

Cut the dendrogram/tree into two main groupl/clusters:

```{r}
grps <-  cutree(wisc.pr.hclust, k=2)
table(grps)
```


I want to know how the clustering in `grps` with values of 1 or 2 correspond the expert `diagnosis`.
```{r}
table(grps, diagnosis)
```

My clustering **groups 1** are mostly "M" diagnosis (179) and my clustering **group 2** are mostly "B" diagnosis

24 FP
179 TP
333 TN
33 FN

Sensitivity TP/(TP+FN)

```{r}
(179/(179+33))
```

Sensitivity TN/(TN+FP)

```{r}
333/(333+24)
```

> Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

There is a high specificity  and strong sensitivity compared to clustering on the original data. 

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results? - We should prioritize the patient with cluster 1 for a follow up because they are maglignant. 

