---
title: "Class 12: Genome informatics"
author: "Jolie Adair (PID 17337497)"
format: pdf
---

## Section 4: Population Scale Analysis [HOMEWORK] 

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.

This is the final file you got ( https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt ). The first column is sample name, the second column is genotype and the third column are the expression values. 

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. Hint: The read.table(), summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein.  Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object. There is also the medium() and summary() function that you can use to check your understanding. 

The dataset included 462 total samples, with genotype counts of 108 A/A, 233 A/G, and 121 G/G. The median expression levels were 31.25 for A/A, 25.06 for A/G, and 20.07 for G/G. These results indicate that ORMDL3 expression is highest in individuals with the A/A genotype and lowest in those with the G/G genotype.

```{r}
bob <- read.table("BunchOfGenes")
head(bob)

```
```{r}
nrow(bob)
```

```{r}
table(bob$geno)
```


```{r}
tapply(bob$exp, bob$geno, median)
```

```{r}
summary(bob)
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

The boxplot demonstrates that individuals with the A/A genotype exhibit the highest levels of ORMDL3 expression, whereas those with the G/G genotype show the lowest levels. This pattern suggests that the rs8067378 SNP is linked to variation in ORMDL3 expression, indicating that this SNP likely influences gene expression.

```{r}
library(ggplot2)


ggplot(bob) +
aes(geno, exp, fill = geno) +
geom_boxplot(notch = TRUE)
```

