---
title: "Class 18: Pertussis and the CMI-PB project"
author: "Jolie Adair (PID: A17337497)"
format: pdf
toc: TRUE
---

## Background 
Pertussis (whooping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone but is most deadly for infants (under 1 year of age).


# CDC Tracking Data
The CDC data 

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(datapasta)

cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L,2024L, 2025L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,21996,22538 )
       )
```

> Q. I want a plot of year vs cases

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(x = Year, y = Cases) +
  geom_point() +
  geom_line() +
  labs(title = "CDC Pertussis Cases Over Time",
       x = "Year",
       y = "Number of Cases")
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

After the introduction of the wP vaccine in 1946, pertussis cases dropped dramatically, but cases began to increase again after the switch to the aP vaccine in 1996.

```{r}
ggplot(cdc) +
  aes(x = Year, y = Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col="orange", lty=2) +
  geom_vline(xintercept = 1996, col="red", lty =2) +
  geom_vline(xintercept = 2020, col="blue", lty =2) +
  labs(title = "CDC Pertussis Cases Over Time",
       x = "Year",
       y = "Number of Cases")
```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the aP vaccine was introduced, pertussis cases gradually increased, possibly because immunity from the acellular vaccine weakens faster than the original whole-cell vaccine.

## Exploring CMI-PB data 

The CMI-PB project < https://www.cmi-pb.org/ >:      The mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, make available a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection.

They make their data available as JSON format API. We can read this into R with the `read_json()` function from th **jsonlite** package.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject)
```

>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

87

```{r}
table(subject$infancy_vac)
```


> Q5. How many Male and Female subjects/patients are in the dataset?

way more females at 112 than males at 60

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

The dataset is mostly composed of White and Asian participants, with 48 White females and 32 White males, and 32 Asian females and 12 Asian males, while other racial groups are represented in much smaller numbers.

>Q. is this repersentative of the US population?

No, it is not representative of the U.S. population because the dataset is dominated by White and Asian participants while other racial groups are underrepresented.



## Side-Note: Working with dates

```{r}
library(lubridate)
```


```{r}
today()
```
```{r}
today() - ymd("2000-01-01")
```
```{r}
time_length(today() - ymd("2000-01-01"), "years")
```
> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
time_length(mean(subject$age[subject$infancy_vac == "wP"]), "years")
```
```{r}
time_length(mean(subject$age[subject$infancy_vac == "aP"]), "years")
```
The average age of wP individuals is about 36.8 years while the average age of aP individuals is about 28.1 years, showing that wP individuals are older because the whole-cell vaccine was used earlier.

> Q8. Determine the age of all individuals at time of boost?

The ages at the time of booster vaccination vary across individuals but generally range from about 18 to about 50 years old.
```{r}
subject$age_at_boost <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
```

```{r}
time_length(subject$age_at_boost, "years")
```


> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Yes, the two groups appear significantly different because the wP group is generally older while the aP group is mostly younger.

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill = as.factor(infancy_vac)) +
  geom_histogram(show.legend = FALSE) +
  facet_wrap(vars(infancy_vac), nrow = 2) +
  xlab("Age in years")
```



## Joining multiple tables


```{r}
table(subject$race, subject$biological_sex)
```

We can read more tables from the CMI-PB database

```{r}
speciman <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = T)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = T)
```

```{r}
head(speciman)
```
```{r}
head(ab_titer)
```
To make sense of all this data we need to "join" (a.k.a."merge" or "link") all these tables together. only then will uou know that a give Ab measurment was collected on a ceretin data form a certion data from a certin wP or aP subject (from the `subject` table)

we can use **dplyer** and the `*_join()` family of functions to do this.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

Yes, the two groups appear significantly different because the wP group is generally older while the aP group is mostly younger.

```{r}
library(dplyr)
meta <- inner_join(subject, speciman)
head(meta)
```


let's do one more `inner_join()` to join the `ab_titer` which all out `meta` data

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

Six antibody isotypes are measured in the dataset: IgE (6698), IgG (7265), IgG1 (11993), IgG2 (12000), IgG3 (12000), and IgG4 (12000).

```{r}
table(abdata$isotype)
```
>Q hoe many diffrent `antigen' values are measured?

There are 16 different antigen values measured in the dataset.

```{r}
table(abdata$antigen)
```


> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

There are four dataset values: 2020_dataset, 2021_dataset, 2022_dataset, and 2023_dataset. The most recent datasets contain fewer rows than the earlier datasets, likely because fewer samples have been collected or processed so far.

```{r}
table(abdata$dataset)
```

Lets focus on IgG istype

```{r}
igg <- abdata |>
  filter(isotype=="IgG")
head(igg)
```

We make a plot of `MFI_normalised` values fpr all `antigen` vlaues

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

The antigens PT, FIM2/3, and FHA appear to have the widest range of antibody titer values.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

The antigens "PT", "FIM2/3" and "FHA" apeaer to have the widest range of values

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens PT, FHA, FIM2/3, and PRN show the largest differences in IgG antibody titers. These antigens are components of the pertussis vaccine, so they trigger stronger immune responses, while unrelated antigens such as OVA or measles show much smaller changes.


> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

Two antigens were selected for analysis: PT, a pertussis toxin antigen found in the vaccine, and OVA, which serves as a control antigen not related to pertussis.

```{r}
pt_ova <- igg |>
  filter(antigen %in% c("PT", "OVA"))
```

```{r}
ggplot(pt_ova) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

For this comparison I selected the control antigen OVA and the pertussis antigen PT.
OVA is not part of the pertussis vaccine and serves as a baseline control, while PT
(Pertussis toxin) is a major antigen in the vaccine and should show a stronger
immune response.

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

The PT antigen shows a clear increase in IgG antibody levels after the booster vaccination, while the control antigen OVA shows little change over time because it is not related to the pertussis vaccine.

>Q17. Do you see any clear difference in aP vs. wP responses?

Yes, there appears to be some differences between aP and wP responses, with wP individuals generally showing slightly stronger antibody responses to several pertussis antigens compared to aP individuals.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```
```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen,) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```


Is there a difference with time (I.e. before booster shot vs. after booster shot)

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>%
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

> Q18. Does this trend look similar for the 2020 dataset?

Yes, the trend looks similar in the 2020 dataset. IgG antibody levels against PT increase after the booster vaccination, peak around day 14, and then gradually decline, which is consistent with a typical immune response.

