---
title: "Class 7: Machine Learning"
author: "Jolie (PID A17337497)"
format: pdf
---

## Background
Today we begin our exploration of important machine learning methods with a focus on **clustering** and **dimensionallity reduction**.

To start testing these methods let's make up some smaple data to cluster where we know what the answer should be.


```{r}
hist( rnorm(3000, mean =10) )
```
> Q. an you generate 30 numbers centered at +3 and 30 numbers at -3 taken at random from a noormal distrubition?

```{r}
tmp <- c(rnorm(30, mean=3), 
         rnorm(30, mean= -3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

## K-means clustering

The main function in "base R" for K-means clustering is called `kmean()`, let's try it out:

```{r}
k <- kmeans(x, centers=2)
k
```

> Q. What component of your kmeans result objects has the cluster centers?


```{r}
k$centers
```


> Q. What component of your kmeans result object has the cluster size (i.e. the main clustering result: which points are in which clusters)?

```{r}
k$size
```

> Q. What component of your kmeans reults object has the cluster membership vector (i.e. the main clustering result: which points are in which cluster)?

```{r}
k$cluster
```


> Q. Plot the results of clustering (i.e. our data colored by the clustering result) along with the cluster centers.

```{r}
plot(x, col=k$cluster)
points(k$centers, col="green", pch=15, cex=2)
```

> Q. Can you run `kmeans()` again and cluster `x` into 4 clusters and plot the results just like we did above with coloring by cluster and the cluster centers shown in blue?


```{r}
K4 <- kmeans(x, centers=4)
K4
plot(x, col=K4$cluster)
points(K4$centers, col="green", pch=15, cex=2)

```

> **Key-Point** Kmeans will always return the clustering that we asked for (this is the "K" or "centers" in Kmeans)!

```{r}
k$tot.withinss
```


## Hierarchical clustering

The main function for hierarchical clustering in base R is called `hclust()`. One of the main differences with respect to the `kmeans()` function is that you can not just pass your input data directly to `hclust()` - it needs a "distance matrix" as input. We can get this from lots of places including the `dist()` function.

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can "cut" the dendrogram or "tree" at a given height to yield our "clusters". For this we use the function `cutree()`

```{r}
plot(hc)
abline(h=10, col="magenta")
grps <- cutree(hc, h=10)
```

```{r}
grps
```

> Q. Plot our data `x` colored by the clustering results from `hclust()` and `cutree()`?

```{r}
grps <- cutree(hc, h=10)
plot(x, col=grps)

```

## Principal Component Analysis (PCA)

PCA is a popular dimensionality reduction technique that is widely used for bioinformatics

## PCA of UK food data

Read data on food consumption in the UK


```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```
There are 17 rows and 5 columns

 
It looks like the row names are not set properly. We can fix this.

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?


```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```

A better way to do this is fix the row names assignments at import time:

```{r}
x <- read.csv(url, row.names=1)
```




> Q3: Changing what optional argument in the above barplot() function results in the following plot?

Changing the beside argument to false.

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```
Skip ggplot Q4

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

If a given point lies on the diagonal in the plot then that mean that the relationship between the two countries on the graph shares a similar value, if not a symmetrical than that means there is 

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```


### Heatmap

We can install the **pheatmap** package with the `instal.packages()` command that we used previously. Remember that we always run this in console and not a code chunk in our quarto document.

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

Of all these plot really only the `pairs()` plot was useful. This however took a bit of work to interpret and will it scale when I am looking at bigger datasets.

## PCA the rescue

The main function in 'base R' for PCA is called `prcomp()`

```{r}
pca <- prcomp(t(x))
summary(pca)
```
PC1 cum port + PC2 cum port = PC3 cum port


> Q. How much variance is captured in the first PC?

67.4%

> Q. How many PCs do I need to cpature at leats 90% of the total variance in the dataset?

Two PCs capture 96.5% of the total variance.

> Q. Plot our main PCA result. Folks can tell this different things depending on their fied of study e.g. "PC plot", "ordienation plot", "Score plot", "PC1 vs PC2 plot" ...

```{r}
attributes(pca)
```

To generate our PCA score plot we want the `pca$x` component of the resulting object

```{r}
pca$x
```

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```
8. 
```{r}
library(ggplot2)

ggplot(pca$x) + aes(PC1, PC2) + geom_point(col=my_cols)
  

```

## Digging deeper (variables loadings)

How do the original variable (i.e. the 17 different foods) contribute to our new PCs?

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```



